---
title: "Federalist Disputed Essays Analysis - Cluster Analysis"
author: "Bharath Karumudi"
date: "6/3/2019"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction
The Federalist Paper data set (Disputed_Essay_data.CSV) is provided. The features are a set of “function words”, for example, “upon”. The feature value is the percentage of the word occurrence in an essay. For example, for the essay “Hamilton_fed_31.txt”, if the function word “upon” appeared 3 times, and the total number of words in this essay is 1000, the feature value is 3/1000=0.3%

### Objective
To find the Author for the disputed essays using Cluster Analysis: K-means and Hierarchical Clustering algorithms.

### Install Packages
```{r install_packages, results='hide'}
#install.packages("dplyr")
#install.packages("ggplot2")
#install.packages("factoextra")
```

### Load Libraries

```{r load_libs, results='hide'}
library(dplyr)
library(ggplot2)
library(factoextra)
```


### Load the dataset

```{r load_dataset}
disputed_data <- (read.csv("Disputed-Essays-ClusterAnalysis_files/Disputed_Essay_data.csv"))
#Lets see the data:
head(disputed_data)
```

### Data Cleansing
```{r data_cleansing}

filtered_dataset <- filter(disputed_data, author != "Jay" & author != "HM")

#Remove the Author column
filtered_data_noauth<- select(filtered_dataset, -c(1))

#Add File names to the row names in the dataset
rownames(filtered_data_noauth) <- filtered_data_noauth[,1]

cleansed_data <- select(filtered_data_noauth, -c(1))

cleansed_data <- na.omit(cleansed_data)
cleansed_data <- scale(cleansed_data)

#Lets see the cleansed data:
head(cleansed_data)

```

### Building Cluster Analysis using K-Means
```{r cluster_analysis}

distance <- get_dist(cleansed_data, method = "euclidean")
fviz_dist(distance, gradient = list(low = "#00AFBB", mid = "white", high = "#FC4E07"))
km_output <- kmeans(cleansed_data, centers = 2, nstart = 25, iter.max = 100, algorithm = "Hartigan-Wong")
fviz_cluster(km_output, data = cleansed_data)

```

From above we can conclude that the disputed essays were written by *Madison*.

### Building Cluster using HAC (Hierarchical Clustering)

```{r hac}

hac_output <- hclust(dist(cleansed_data, method = "euclidean"), method = "complete")
plot(hac_output)

hac_cut <- cutree(hac_output, 1)

table(hac_cut, filtered_dataset$author)

```

From above results, we can see the model has predicted that the disputed essays are written by *Madison*.


### Conclusion
From both the models- K-Means and HAC, we can see the disputed Federal essays are written by *Madison*.

### References

1. Euclidean Distance
2. Hierarchical clustering documentation
